---
title: "A gentle intro to DEB theory and ontogenetic growth with ODEs in R"
author: "James Maino"
date: "22 April 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
```

What I want to do in the following is introduce some basic Dynamic Energy Budget (DEB) growth equations in the form of ordinary differential equations (ODEs). First, we set values for some of the core DEB parameters, which we have hopefully discussed. If not, do so now...

```{r}
# initial vars
V0 = 0.01 # intial structure, cm^3

# set parameters
p_Am =  100   # SA specific assimilation rate, J/cm^2/d
E_G   = 28   # cost of growth, J/cm^3
v     = 0.02 # conductance, cm/d
p_M   = 18   # vol. specific maintenance rate, J/cm^3/d
f     = 1    # scaled functional response of feeding, -
```


# Setting up the ODEs
We are using the differential equation solver function `ode()` in the R package `deSolve::`. This function takes four necessary arguments: 

`y`:	the initial (state) values for the ODE system, a vector. If y has a name attribute, the names will be used to label the output matrix.

`times`: time sequence for which output is wanted; the first value of times must be the initial time.

`func`: either an R-function that computes the values of the derivatives in the ODE system (the model definition) at time t, or a character string giving the name of a compiled function in a dynamically loaded shared library.
The return value of func should be a list, whose first element is a vector containing the derivatives of y with respect to time, and whose next elements are global values that are required at each point in times. The derivatives must be specified in the same order as the state variables y.

`parms`: parameters passed to func.

### 1. Initialise state variables
```{r}
state <- c(res = f*p_Am/v*V0,
           V   = V0)
```

### 2. Set integration time points
```{r}
times <- seq(0,5000,length = 1000)
```

### 3. Create function containing ODEs
```{r}
growth<-function(t, state, pars) {
  # parameters
  p_Am= pars[1]
  E_G = pars[2]
  v   = pars[3]
  p_M = pars[4]
  f   = pars[5]
  
  
  # state variables
  res = state[1]
  V   = state[2]
  
  # rate of change
  dres = f*p_Am*V^(2/3) - res* (E_G*v/V^(1/3) + p_M)/(res/V + E_G);
  dV   = (res*v/V^(1/3) - p_M*V)/(res/V + E_G);
  
  # return the rate of change
  list(c(dres, dV))
}
```

### 4. Wrap parameters to pass to function
```{r}
pars       <- c(p_Am  = p_Am,
                E_G   = E_G,
                v     = v,
                p_M   = p_M,
                f     = f)
```

## Solve ODEs
Now it is a simple matter of running the function.
```{r}
resV <- ode(y = state, times = times, func = growth, parms = pars)

```

# Plot results
Here we plot the output of the state variables reserve and structure through time. These state variables determine the state of the system, irrespective of how we got there. From these state variables we can calculate a range of other biological variables, such as total biomass, assimmilation rate, growth rate, mainentance rate, etc. Importantantly, this may be data we actaully have measured on species and how we go about fitting a DEB model to data. In this example we only looked at the non-reproductive growth phase, but the standard DEB model also includes embryonic, and reproductive phases, for the cost of only a few more parameters $\kappa$, $\dot{k_J}$, $E_{Hb}$, $E_{Hj}$. 
```{r}
plot(times, resV[,3],'l', xlab = 'time, d', ylab = expression(structure*","~ cm^3))
plot(times, resV[,2]/resV[,3],'l', xlab = 'time, d', ylab = expression(reserve/structure*","~J/cm^3))
```

## Mathematical overkill? 
So far it may seem like DEB theory is a complex presentation of the von Bertalanffy growth curve, which has the simple analytical solution:

$$V = V_\infty(1-(1-V_0^{1/3}/V_\infty^{1/3})e^{-at})^3$$

Indeed, plotting the analytical solution (black line) over the numerical solution (red circles) shows they are equivilent.


```{r}
V_inf = (p_Am/p_M)^3
a<-1/(3/p_M/E_G + 3*V_inf^(1/3)/v)

plot(times, V_inf*(1-(1-V0^(1/3)/V_inf^(1/3))*exp(-a*times))^3, 'l', ylab = expression(V*','~cm^3));points(resV[seq(1,length(times),100),1],  resV[seq(1,length(times),100),3], col = 'red')
```

But the ODE method also offers a simple solution to non-steady state conditions, such as non-constant food or temperature.

For example, say that food availability varied throught time as:

```{r}
plot(times,(0.1*sin(times*2*pi/365)+0.8), 'l', ylab = 'feeding response, -', main = 'f = 0.1*sin(times*2*pi/365)+0.8')
```

In this example you could probably use some fancy calculus to find another analytical solution, but it is straight forward to solve numerically after modifying the growth equation to capture the variance in food conditions.

```{r}
growth2<-function(t, state, pars) {
  # parameters
  p_Am= pars[1]
  E_G = pars[2]
  v   = pars[3]
  p_M = pars[4]
  f   = (0.1*sin(t*2*pi/365)+0.8) # change the functional response.
  
  # state variables
  res = state[1]
  V   = state[2]
  
  # rate of change
  dres = f*p_Am*V^(2/3) - res* (E_G*v/V^(1/3) + p_M)/(res/V + E_G);
  dV   = (res*v/V^(1/3) - p_M*V)/(res/V + E_G);
  
  # return the rate of change
  list(c(dres, dV))
}
```


Solve ODEs for varying food
```{r}
resV2 <- ode(y = state, times = times, func = growth2, parms = pars)

```

Plot results for varying food
```{r}
plot(times, resV2[,3],'l', col = 'red', ylim =c(0,200), xlab = 'time, d', ylab = expression(structure*","~ cm^3));lines(times, resV[,3])
plot(times, resV2[,2]/resV2[,3],'l', col = 'red', xlab = 'time, d', ylab = expression(reserve/structure*","~J/cm^3));lines(times, resV[,2]/resV[,3])
```

A key achievement of DEB theory is the insight into the mechanistic basis of well known growth curves, and their extension to non-steady state conditions and other life-history stages. It can help to explain indirect calorimetry (energy expenditure from gas exchange), and contraints on energy flows within and between species. It potentially serves as a framework for the quantitative intrepretation of the life-histories, metabolic architecture, and energy budgets of diverse species in terms of variance in their parameter values. 
